---
title: "Mesocosm model in Fortran"
author: "Dick and Karline"
date: "27/05/2021"
output: pdf_document
---

# Packages

```{r, echo=FALSE, message = FALSE}
require(FME)
require(seacarb)
require(marelac)
```

# Accessory functions

## The 13C conversion functions

```{r}
Cstd  <- 0.0112372

dC_RC <- function(dC){signif(((dC/1000)+1)*Cstd, 8)}

dC_FC <- function(dC){as.double(dC_RC(dC) / (dC_RC(dC) + 1))}
```

# Read data

```{r}
# for forcing functions
pH   <- readRDS("pH.rds")

Temp <- readRDS("Temp.rds")
Temp <- rbind(Temp, cbind(time = Temp[,1]+8,  Temp=Temp[,2]))
Temp <- rbind(Temp, cbind(time = Temp[,1]+16, Temp=Temp[,2]))

PAR  <- readRDS("PAR.rds")
PAR  <- rbind(PAR, cbind(time = PAR[,1]+8,  PAR=PAR[,2]))
PAR  <- rbind(PAR, cbind(time = PAR[,1]+16, PAR=PAR[,2]))

# observed data
oxysat    <- readRDS("oxysat.rds")
Data_d13C <- readRDS("Data_d13C.rds")
Data_13C  <- readRDS("Data_13C.rds")
Data_wc   <- readRDS("Data_wc.rds")
Data_C    <- readRDS("Data_C.rds")

```

# Create forcing functions

```{r, echo=FALSE}
#par, temp, pH, oxysat, K0, K1, K2
times.default <- seq(from=0, to=32, length.out=321)

PAR.default    <- cbind(times.default, approx(PAR,  xout = times.default, rule = 2)$y)
temp.default   <- cbind(times.default, approx(Temp, xout = times.default, rule = 2)$y)
pH.default     <- cbind(times.default, approx(pH,   xout = times.default, rule = 2)$y)
oxysat.default <- cbind(times.default, gas_O2sat(S=32, temp.default[,2])/32*1000)
k0.default     <- cbind(times.default, as.double(K0(S=32, T=temp.default[,2], P=0, Patm=1, warn="y")))
k1.default     <- cbind(times.default, as.double(K1(S=32, T=temp.default[,2], P=0)))
k2.default     <- cbind(times.default, as.double(K2(S=32, T=temp.default[,2], P=0)))

```

# default parameter values

```{r}
pars.default <-   
	   c(DIC_dC_b =  2050.0, 
	  	H         =     1.    , # height of the tank [m]
	   	Kexc_DIC  =     5.0   , # Gas exchange coeff dic [umol C l-1 uatm-1 d-1] VdM04
  	  pCO2atm   =   400.0   , # partial pressure CO2 in air
	   	Kexc_Oxy  =     4.8   , # Gas exchange coeff oxygen [±20 cm h-1 => 20 /100*24 = 4/8 m d-1] Kihm
	  	Q10       =     2.    , # temp limitation of process rates [-]
 
	  	ON_nit    =     2.0   , # O : N ratio in nitrification
	  	nitr_NH4  =     0.03  , # nitrification rate (d-1)
	  	
	  	alpha     =     0.025 , # slope of the P-I curve [umol m-2 s-1 d-1] VdM04
    	K_DIC_P1  =   250.0   , # half-saturation constant DIC uptake [mmol C m-3] 
      K_NH4_P1  =     0.5   , # half-saturation constant NH4 uptake [mmol N m-3] 
	  	K_NOX_P1  =     0.5   , # half-saturation constant NOX uptake [mmol N m-3] 
	  	psi_P1    =     1.5   , # NH3 inhibition for NOX uptake [m3 mmol N-1]
	  	uN_P1     =     2.0   , # net growth rate for NOX and NH4 [mmol N mmol C-1 d-1]
	  	uC_P1     =     5.5   , # net growth rate [d-1]
	  	minCN_P1  =     6.0   , # minimum CN ratio [mmol C mmol N-1]
	  	maxCN_P1  =    20.0   , # maximum CN ratio [mmol C mmol N-1]
      gamma1_P1 =     0.02  , # leakage fraction [-]
      gamma3_P1 =     0.05  , # DOC produced due to Nlim [-]
      resb_P1_C =     0.01  , # basal respiration [d-1]
      resa_P1_C =     0.10  , # activity respiration [-]
      mort_P1_N  =     0.045 , # mortality rate [d-1] 

    	K_DIC_P2  =   250.0   , # half-saturation constant DIC uptake [mmol C m-3] 
      K_NH4_P2  =     0.5   , # half-saturation constant NH4 uptake [mmol N m-3] 
	  	K_NOX_P2  =     0.5   , # half-saturation constant NOX uptake [mmol N m-3] 
	  	psi_P2    =     1.5   , # NH3 inhibition for NOX uptake [m3 mmol N-1]
	  	uN_P2     =     2.0   , # net growth rate phyto1 for NOX and NH4 [mmol N mmol C-1 d-1]
	  	uC_P2     =     5.5   , # net growth rate phyto1 [d-1]
	  	minCN_P2  =     6.0   , # minimum CN ratio [mmol C mmol N-1]
	  	maxCN_P2  =    20.0   , # maximum CN ratio [mmol C mmol N-1]
      gamma1_P2 =     0.02  , # leakage fraction [-]
      gamma3_P2 =     0.05  , # DOC produced due to Nlim [-]
      resb_P2_C =     0.01  , # basal respiration [d-1]
      resa_P2_C =     0.10  , # activity respiration [-]
      mort_P2_N =     0.045 , # mortality rate [d-1] 
	  	
	  	delta     =     0.65  , # part of mortality and grazing flux to DMl (1-delta to DMs)
	  	epsilon   =     0.34  , # part of mortality and grazing flux to DM (1-epsilon to PM)
	  	
      CN_B      =     5.1   , # molar C:N ratio of bacteria
	  	uB        =    13.3   , # max uptake of DMl by Bacteria, d-1 {VdM04}
	  	K_DMl_B   =    25.0   , # half-saturation uptake of DMl by Bacteria [mmolC m-3], {VdM04}
      K_NH4_B   =     0.5   , # half-saturation uptake of NH4 by Bacteria [mmolN m-3], {VdM04}
      omega     =     0.19  , # bacterial growth efficiency [-]
	  	mort_B    =     0.05  , # bacterial mortality (quadratic term) [mmol C-1 d-1], {VdM04}
	  	
  	  maxCN_PMw =    15.0   , # max molar CN ratio of PMw [C/N]
      deg_PMw_C =     0.29  , # degradation rate of C-detritus into DMl + DMs [d-1], C component more refractory
      deg_PMw_N =     0.33  , # degradation rate of N-detritus into DMl + DMs [d-1], N component less refractory
	  	
  	  uDMs      =     4.0   , # max dissolution rate of DMs_C [d-1]  
      K_DMs     =   417.0   , # half-saturation constant for dissolution  of DMs_C [mmolC m-3]
      
	  	w         =     0.05  , # settling velocity of PMw [m/d] [default = 0.5]
	  	
	  	maxCN_PMs =    15.0   , # max molar CN ratio of PMs [C/N]
	  	deg_PMs_C =     0.29  , # degradation rate of C-detritus into DMl + DMs [d-1], C component more refractory
      deg_PMs_N =     0.33  , # degradation rate of N-detritus into DMl + DMs [d-1], N component less refractory
      
	  	CN_mZP    =     6.25  , # molar CN ratio zooplankton
      gra_mZP   =     1.0   , # max grazing mZP [d-1]
      K_mZP     =     1.0   , # half-saturation grazgin by mZP [mmolN m-3]
      pP1_mZP   =     0.4   , # preference for P1 by mZP [-]
      pP2_mZP   =     0.3   , # preference for P2 by mZP [-]
      pB_mZP    =     0.2   , # preference for B by mZP [-] 
      pPMw_mZP  =     0.1   , # preference for PMw by mZP [-]
      phi       =     0.23  , # mZP feeding losses to DOM [-]
      beta_mZP  =     0.75  , # ass eff mZP [-]
      resb_mZP_C=     0.03  , # basal resp mZP [d-1]
      resa_mZP_C=     0.25  , # fraction of assimilate food that is respired by mZP [-]
      mort_mZP  =     0.3   , # zooplankton mortality [d-1]
	  	K_mor_mZP =     0.2   , # half-saturation constant for zooplankton mortality [mmolN m-3]

	  	CN_Oy     =     6.25  , # CN ratio oysters [C/N}]
	   	filt_Oy   =     0.002 , # filtration rate oysters [m^3 mmol C-1 d-1]
	  	fae_Oy    =     0.0   , # fraction released as faeces
	  	pse_Oy    =     0.65  , # fraction released as pseudeo-faeces
      resa_Oy_C =     0.25  , # fraction of assimilated food that is respired by Oy [-]
      resb_Oy_C =    0.005    # maint resp Oy [d-1]
      ) 
```


## Initial conditions

```{r}
Yini.meso <- function(pars, tempini = 21){
 Oxy_i      <-  gas_O2sat(S = 32, t = tempini) /32*1000*1.1

NOX_i      <-     1.1                      # start NOX conc [mmolN m-3]
NH4_i      <-     2.0                      # start NH4 conc [mmolN m-3]
	
DIC_i      <-  2200.0                      # start DIC concentration [mmolC m-3]
DIC_dC_b   <-  2050.0
DIC_13C_i  <-  dC_FC(DIC_dC_b) * DIC_i

P1_N_i     <-     0.75                     # start N conc [mmolN m-3], P1 = diatoms
P1_C_i     <-     7.5                      # start C conc [mmolC m-3], P1 = diatoms
P1_dC_b    <-   -20.0                      # d13C bckgrnd
P1_13C_i   <-  dC_FC(P1_dC_b) * P1_C_i     # start 13C conc [mmolC m-3]

P2_N_i     <-     0.3                      # start N conc [mmolN m-3], P2 = other phyto
P2_C_i     <-     3.0                      # start C conc [mmolC m-3], P2 = other phyto
P2_dC_b    <-   -21.0                      # d13C bckgrnd
P2_13C_i   <-  dC_FC(P2_dC_b) * P2_C_i     # start 13C conc [mmolC m-3]

DMl_N_i     <-     1.0                     # start N conc [mmolN m-3], DMl = labile dissolved organic matter
DMl_C_i     <-    10.0                     # start C conc [mmolC m-3], DMl = labile dissolved organic matter
DMl_dC_b    <-   -20.0                     # d13C bckgrnd
DMl_13C_i   <-  dC_FC(DMl_dC_b) * DMl_C_i  # start 13C conc [mmol m-3]

DMs_N_i     <-    19.0                     # start N conc [mmol m-3], DMs = semi-labile dissolved organic matter
DMs_C_i     <-   190.0                     # start C conc [mmol m-3], DMs = semi-labile dissolved organic matter
DMs_dC_b    <-   -20.0                     # d13C bckgrnd
DMs_13C_i   <-  dC_FC(DMs_dC_b) * DMs_C_i  # start 13C conc [mmol m-3]

PMw_N_i     <-     4.3                     # start N conc [mmol m-3], PMw_N = part. OM in water column
PMw_C_i     <-    43.                      # start C conc [mmol m-3], PMw_C = part. OM in water column
PMw_dC_b    <-   -20.0                     # d13C bckgrnd
PMw_13C_i   <-  dC_FC(PMw_dC_b) * PMw_C_i  # start 13C conc [mmol m-3]

PMs_N_i     <-     0.001                   # start N conc [mmol m-3], PMs = part. OM on the sediment (tank floor)
PMs_C_i     <-     0.006                   # start C conc [mmol m-3], PMs = part. OM on the sediment (tank floor)
PMs_dC_b    <-   -20.0                     # d13C bckgrnd
PMs_13C_i   <-  dC_FC(PMs_dC_b) * PMs_C_i  # start 13C conc [mmol m-3]

B_C_i       <-     6.0                     # start C conc [mmol m-3], B = bacteria
B_N_i       <- B_C_i/as.numeric(pars["CN_B"])# start N conc [mmol m-3], B = bacteria
B_dC_b      <-   -20.0                     # d13C bckgrnd
B_13C_i     <-  dC_FC(B_dC_b) * B_C_i      # start 13C conc [mmol m-3]

mZP_C_i     <-     2.                      # start C conc [mmol m-3]
mZP_N_i     <- mZP_C_i/as.numeric(pars["CN_mZP"]) # start N conc [mmol m-3], mZP = mesozooplankton
mZP_dC_b    <-   -20.0                     # d13C bckgrnd
mZP_13C_i   <-  dC_FC(mZP_dC_b) * mZP_C_i  # start 13C conc [mmol m-3]

Oy_C_i      <-    1e-6                     # start C conc [mmol m-2], note m-2
Oy_N_i      <- Oy_C_i/as.numeric(pars["CN_Oy"])# start N conc [mmol m-2], note m-2, Oy = Oysters
Oy_dC_b     <-   -21.0                     # d13C bckgrnd
Oy_13C_i    <-  dC_FC(Oy_dC_b) * Oy_C_i    # start 13C conc [mmolC m-2]

ini   <- 
	  c(Oxy      =      Oxy_i     ,
	  	NOX      =      NOX_i     ,
	  	NH4      =      NH4_i     ,
	  	DIC      =      DIC_i     ,          
	  	DIC_13C  =  DIC_13C_i     ,
			P1_N     =     P1_N_i     ,   
			P1_C     =     P1_C_i     ,   
      P1_13C   =   P1_13C_i     ,
			P2_N     =     P2_N_i     , 
			P2_C     =     P2_C_i     , 
	  	P2_13C   =   P2_13C_i     , 
	  	DMl_N    =    DMl_N_i     ,
	  	DMl_C    =    DMl_C_i     ,
	  	DMl_13C  =  DMl_13C_i     ,
	  	DMs_N    =    DMs_N_i     ,
	  	DMs_C    =    DMs_C_i     ,
	  	DMs_13C  =  DMs_13C_i     ,
	  	PMw_N    =    PMw_N_i     ,
	  	PMw_C    =    PMw_C_i     ,
	  	PMw_13C  =  PMw_13C_i     ,
	  	PMs_N    =    PMs_N_i     ,
	  	PMs_C    =    PMs_C_i     ,
	  	PMs_13C  =  PMs_13C_i     ,
	  	B_N      =      B_N_i     ,
	  	B_C      =      B_C_i     ,
	  	B_13C    =    B_13C_i     ,
			#uZ_c    =     
			mZP_N    =    mZP_N_i     ,
			mZP_C    =    mZP_C_i     ,
	  	mZP_13C  =  mZP_13C_i     ,
			Oy_N     =     Oy_N_i     ,
			Oy_C     =     Oy_C_i     ,
	  	Oy_13C   =   Oy_13C_i     ,
	  	MB_N     = 0              ,
			MB_C     = 0              ,
	  	MB_13C   = 0
	  	)
 return(ini)
}

ini <- Yini.meso(pars.default, temp.default[1,2])
```

## Events
```{r}
oy_C <- c(8, 22, 47)
oy_N <- oy_C / as.numeric(pars.default["CN_Oy"])

Oy_dC_b     <-   -21.0                     # d13C bckgrnd
oy_13C <- dC_FC(Oy_dC_b) * oy_C
# experiment 2, 3, and 4

oy_C   <- c(8, 22, 47)
oy_N   <- oy_C / as.numeric(pars.default["CN_Oy"])
oy_13C <- dC_FC(Oy_dC_b) * oy_C

ToChange <- c("Oxy"    , "NOX"    , "NH4"    , "DIC"    , "DIC_13C" ,
	 	 			  "P1_N"   , "P1_C"   , "P1_13C" , "P2_N"   , "P2_C"    , "P2_13C"  ,
						"DMl_N"  , "DMl_C"  , "DMl_13C", "DMs_N"  , "DMs_C"   , "DMs_13C" ,
						"PMw_N"  , "PMw_C"  , "PMw_13C", "PMs_N"  , "PMs_C"   , "PMs_13C" , 
						"B_N"    , "B_C"    , "B_13C"  , "mZP_N"  , "mZP_C"   , "mZP_13C" ,
	          "Oy_N"   ,  "Oy_C"  , "Oy_13C")

Oysterevents <- data.frame(
  var   = rep(ToChange, times = 3),
  time  = c(rep(8, 32), rep(16, 32), rep(24, 32)),
	value = c(ini[1:29], oy_N[1] , oy_C[1] , oy_13C[1] ,   # first the non-oyster states, then states
						ini[1:29], oy_N[2] , oy_C[2] , oy_13C[2] , 
						ini[1:29], oy_N[3] , oy_C[3] , oy_13C[3] ),
						method= "replace")
```



# The fortran DLL

The fortran code is in file meso.f95. 

This needs to be compiled only if the fortran code has changed.

```{r, eval = FALSE}
system("R CMD SHLIB meso.f95")
```

You may need to unload the dll before re-loading it.

```{r}
if (is.loaded("meso.dll")) dyn.unload("meso.dll")
dyn.load(paste("meso", .Platform$dynlib.ext, sep = ""))
```

## Model function

```{r}
RunMesocosm <- function(times = times.default,   # default is from 0 to 32
                        yini = NULL,             # if NULL, the function Yini.meso is called with pars
                        pars = NULL,             # if NULL, the pars.default is used 
                        irradiance=PAR.default, temperature=temp.default,
                        pH=pH.default, oxysat=oxysat.default, 
                        k0=k0.default, k1=k1.default, k2=k2.default, 
                        events = NULL, ...) {
  
  forcings <- list(par=irradiance, temp=temperature, pH=pH, 
                 oxysat=oxysat, K0=k0, K1=k1, K2=k2)


  Parms <- pars.default   # the default parameter values
  nms   <- names(Parms)
  
  Parms[(namc <- names(pars))] <- pars  # overwriting the default paramaeters

  if (length(noNms <- namc[!namc %in% nms]) > 0) 
      warning("unknown names in 'pars': ", paste(noNms, collapse = ", "))

  parms <- unlist(Parms)
  
  if (is.null(yini)) yini <- Yini.meso (parms, tempini = temperature[1,2])
  svarNames <- names(yini)
  
  outnames <- c("oxy_sat", "dC_DIC", "dC_P1", "dC_P2", "dC_DMl", "dC_DMs", "dC_PMw", "dC_PMs", 
							  "dC_B", "dC_mZP", "dC_Oy", "DM_C", "CO2exc", "CO2exc_13C", "pCO2", 
							  "fpar", "ftemp",  "NOXlim_P2", "NH4lim_P2", "gra_Oy_C",
							  "pp_P1_C", "pp_P1_N", "pp_P2_C", "pp_P2_N", "exc_B_N", "exc_mZP_N", "NH4_B_N")

  out <- ode(y = yini, times = times, func = "mesomodel",
    parms = parms, dllname = "meso", initfunc = "initpars", 
    initforc="initforcs", forcings = forcings,  events = events, 
    names = svarNames, outnames = outnames, nout = length(outnames))
  attr(out, "parms") <- parms
  return(out)
}

```

## Solve the model

```{r}
print(system.time(
  out <- RunMesocosm(times=seq(0, 32, length=321), yini=ini,  
                   events=list(data = Oysterevents))
))

out2 <- RunMesocosm(times=seq(0, 32, length=321), yini=ini, pars=list(Kexc_Oxy = 0, uC_P1 = 1.5), 
                   events=list(data = Oysterevents))
```

```{r, fig.width=6, fig.height=6}
plot(out, out2, obs = oxysat, obspar = list(pch = 18, col = "red", cex = 0.5))
```

```{r, fig.width=10, fig.height=10}
plot(out, out2, obs = Data_C, mfrow = c(3,3), 
     obspar = list(pch = 18, cex = 2, col = "red"))
plot(out, out2, obs = Data_d13C, mfrow = c(3,3), 
     obspar = list(pch = 18, cex = 2, col = "red"))

```

# APPENDIX - Fortran code


```{r comment=''}
cat(readLines('meso.f95'), sep = '\n')
```

